A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-10-12 main.tex
4.1, 4.3 edits

We begin by comparing the ordinary Lasso, PC-Lasso, and LMM-Lasso as the level of environmental confounding is varied. Estimation accuracy for Lasso, PC-Lasso, and LMM-Lasso was evaluated based on mean squared error (MSE) and broken down into bias and variance components \pb{let's just report MSE (and maybe MSPE too); breakdown blurs ``bias'' concepts a bit}. Figure \ref{fig:mse} presents these results for all three data types in a 1:1 signal to noise (SNR) setting, where $\eta = 0.5$. The top row of Figure \ref{fig:mse} shows that when $\xi = 0$ (i.e., no environmental confounding), Lasso and LMM-Lasso perform comparably. PC-Lasso results in a slight inflation in MSE compared to the other methods in the simulated data sets due to the fact that 10 additional predictors (PCs) have been added to the model, but none of them actually explain any variability in the outcome.

@@

\begin{figure}[H]
    \centering
    % \includegraphics[scale = 0.9]{figures/mse_diff_subpops}
     \includegraphics[width=0.8\textwidth]{figures/mse_diff_subpops.png}
    \caption{Relative performance of LMM-Lasso, PC-Lasso for fine and coarse subpopulation structures and varying levels of environmental confounding with Independent Subpopulations data, $\eta = 0.5$, and dichotomous-discordant environmental effects. }
    \label{fig:big_vs_small}
\end{figure}

@@

\pb{What do you think about dropping the exponential/dichotomous thing and just going with dichotomous throughout?  Doesn't really seem super important.  Well, we can just keep it in mind and if we need to cut down on space / figures, we can think about it.} \anna{so leave for now?}

\anna{RE-WRITE THIS TO MAKE SURE IT MAKES SENSE WITH NEW FIGURES} The right-hand panels of Figure \ref{fig:big_small_gamma} are consistent with Figure \ref{fig:big_vs_small}, showing superior LMM-Lasso performance in the presence of fine subpopulation structure, regardless of coefficient magnitude or concordance. The upper left hand panel of Figure \ref{fig:big_small_gamma} corresponds to coarse subpopulation structure with discordant coefficients and also appears to be a setting where LMM-Lasso outperforms PC-Lasso, regardless of coefficient magnitude, although there is some overlap with PC-Lasso. The lower left hand panel of Figure \ref{fig:big_small_gamma} shows results for the setting of coarse supopulation structure with concordant coefficients. In this setting, PC-Lasso and LMM-Lasso perform comparably in the presence of exponential-concordant environmental effects, while LMM-Lasso outperforms PC-Lasso in the majority of dichotomous-concordant environmental effects. These findings align with our initial hypothesis that PC-Lasso will perform well in the presence of simple environmental effect patterns that can be well-approximated by a linear combination of the observed data. When that is not the case, either due to the presence of fine subpopulation structure, dichotomous effects, or genetic similarity that does not reflect similar environmental effects, LMM-Lasso exhibits estimation accuracy superior to that of PC-Lasso.

\begin{figure}[H]
    \centering
    % \includegraphics[scale = 0.9]{figures/fig2b.png}
    \includegraphics[scale = 0.9]{figures/mse_diff_hetero.png}
    \caption{Relative performance of LMM-Lasso, PC-Lasso for fine and coarse subpopulation structures and varying environmental effect structures with Independent Subpopulations data, $\eta = 0.5$, and $\xi = 0.8$.}
    \label{fig:big_small_gamma}
\end{figure}

@@

As an additional point of interest, recall that we provide qualitative and quantitative explanations of how the degree to which differential environmental exposures may confound the genotype-phenotype relationship is dependent upon its similarity to population structure. While the objective of these explanations, and this work as a whole, is to elucidate the manner in which failing to correct for unobserved environmental effects can bias SNP effect estimates, equation \eqref{eqn:partition_e} also highlights the manner in which such unobserved and unmodeled environmental effects could lead to substantial bias in narrow-sense heritability estimatess. A more detailed treatment of narrow-sense heritability estimation is beyond the scope of this paper. However, we hope this emphasizes the broad implications and importance of the subtle distinction between the effect of population structure and unobserved environmental heterogeneity.

>>>>

We begin by comparing the ordinary Lasso, PC-Lasso, and LMM-Lasso as the level of environmental confounding is varied. Estimation accuracy for Lasso, PC-Lasso, and LMM-Lasso was evaluated based on mean squared error (MSE) \anna{MSPE sims running - maybe add here}. Figure \ref{fig:mse} presents these results for all three data types in a 1:1 signal to noise (SNR) setting, where $\eta = 0.5$. The top row of Figure \ref{fig:mse} shows that when $\xi = 0$ (i.e., no environmental confounding), Lasso and LMM-Lasso perform comparably. PC-Lasso results in a slight inflation in MSE compared to the other methods in the simulated data sets due to the fact that 10 additional predictors (PCs) have been added to the model, but none of them actually explain any variability in the outcome.

@@

\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.9]{figures/mse_diff_subpops.png}
    \caption{Relative performance of LMM-Lasso, PC-Lasso for fine and coarse subpopulation structures and varying levels of environmental confounding with Independent Subpopulations data, $\eta = 0.5$, and dichotomous-discordant environmental effects. }
    \label{fig:big_vs_small}
\end{figure}

@@

% \pb{What do you think about dropping the exponential/dichotomous thing and just going with dichotomous throughout?  Doesn't really seem super important.  Well, we can just keep it in mind and if we need to cut down on space / figures, we can think about it.} \anna{so leave for now?}

The right-hand panels of Figure \ref{fig:big_small_gamma} are consistent with Figure \ref{fig:big_vs_small}, showing superior LMM-Lasso performance in the presence of fine subpopulation structure, regardless of coefficient magnitude or concordance.

LMM-Lasso appears to perform as well, or better than, PC-Lasso in the presence of coarse subpopulation structure, as shown by the left-hand panels of Figure \ref{fig:big_small_gamma}. The upper left-hand facet in particular shows that PC-Lasso and LMM-Lasso perform more comparably when environmental effect magnitudes are exponential and concordant in nature. 
These findings align with our initial hypothesis that PC-Lasso will perform well in the presence of simple environmental effect patterns that can be well-approximated by a linear combination of the observed data. When that is not the case, either due to the presence of fine subpopulation structure, dichotomous effects, or genetic similarity that does not reflect similar environmental effects, LMM-Lasso exhibits estimation accuracy superior to that of PC-Lasso.

\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.9]{figures/mse_diff_hetero.png}
    \caption{Relative performance of LMM-Lasso, PC-Lasso for fine and coarse subpopulation structures and varying environmental effect structures with Independent Subpopulations data, $\eta = 0.5$, and $\xi = 0.8$.}
    \label{fig:big_small_gamma}
\end{figure}

@@

As an additional point of interest, recall that we provide qualitative and quantitative explanations of how the degree to which differential environmental exposures may confound the genotype-phenotype relationship is dependent upon its similarity to population structure. While the objective of these explanations, and this work as a whole, is to elucidate the manner in which failing to correct for unobserved environmental effects can bias SNP effect estimates, equation \eqref{eqn:partition_e} also highlights the manner in which such unobserved and unmodeled environmental effects could lead to substantial bias in narrow-sense heritability estimates. A more detailed treatment of narrow-sense heritability estimation is beyond the scope of this paper. However, we hope this emphasizes the broad implications and importance of the subtle distinction between the effect of population structure and unobserved environmental heterogeneity.
